\begin{acknowledgements}

Pursuing doctoral studies will have been an epic though challenging adventure, and I am extremely grateful to have spent those years living in the vibrant city of Berkeley, being an active member of UC Berkeley and its EECS department. 

My warmest acknowledgments go to my advisors Nir Yosef and Mike Jordan. Both welcomed me into their group, helped me trust in my work during challenging times, and generally made me a more curious, exigent, and enthusiastic researcher. Thank you, Nir for hiring me into the Berkeley EECS Ph.D. program, sending me to Israel, funding the ``French internship program'' of the Yosef lab, and for relentlessly challenging my ideas until they become better. Thank you Mike for always being a source of inspiration as a researcher, helping me share my time between computational biology and computer science, and for making it possible to intern at Ant Financial as well as Amazon. 

Early in graduate school, I met Jeffrey Regier at Mike's semi-annual pizza party, where we quickly realized our common interest in modeling.  Jeff focused on modeling photon counts for his astronomy work at the time, while I wanted to model transcript counts. Jeff has since been an admirable mentor, teaching me applied Bayesian statistics, software engineering, as well as a great deal of technical writing. Thanks, Jeff for making this Ph.D. a rich and fun experience! 

On my five-year journey, I have had the luck of interacting with many brilliant senior researchers, either inside or outside of the Bay Area. First, I would like to warmly acknowledge Le Song for welcoming me to Hangzhou, and our work on domain adaptation bounds for counterfactual inference. This stay in China will remain a fond memory of my graduate studies. To Inderjit Dhillon for having me at the Berkeley Amazon office, and for the opportunity to learn about eXtreme multi-label classification. To Ido Amit for making me feel like a detective while unraveling the behavior and the localization of immune cells in the MCA205 tumor. To Jennifer Listgarten for providing career perspectives and being present on my qualifying examination committee. To Sandrine Dudoit for inspiring me to apply to Berkeley and for being part of my thesis committee. To Laurent El Ghaoui and Alexandre Bayen for showing me how pleasing and rewarding teaching is, and for helping out with the French Alumni Berkeley association. Many thanks!

I especially acknowledge the staff at UC Berkeley and in particular Shirley Salanio, Kate Marx, Martin Witte, Sanchita Pal. No dissertation would ever get written without their help. 

Outside of Berkeley, I would like to acknowledge the persons who shaped my early research life and encouraged or inspired me to apply to a Ph.D. program. Thanks to Eric Serres, Olivier Krygiel and Roger Savardel from the Lyc\'ee Ferdinand Fabre, and John Bandelier from Kimiyo. Thanks to Erwan Le Pennec and Laurent Massouli\'e for their mentorship at \'Ecole polytechnique. Thanks to Jia Li from Cardiologs for teaching me applied machine learning. Thanks to Jean-Baptiste Michel for inspiring me to get interested in computational biology. And thanks to Allon Klein for welcoming me at the Harvard Medical School for my Master's thesis. As you told me once, I might have caught the biology bug! 

Throughout these years at Berkeley, I am most grateful for sharing my path with many friends. Whether it is about stimulating conversations, getting through the highs and lows of graduate school, pulling all-nighters for conference paper deadlines, reflecting on life in a foreign country and about cultural differences, or simply enjoying a homemade paella, it is always much better done surrounded! I have learned a lot interacting with the members of SAIL, including Ryan Giordano, Ahmed El Alaoui, Nicolas Flammarion, Chi Jin, Lihua Lei, Horia Mania, Philipp Moritz, Robert Nishihara, Jianbo Chen, Nilesh Tripuraneni, Koulik Khamaru, Eric Mazumdar, Alyssa Morrow, Esther Rolf, Aldo Pacchiano, Jake Soloff, Sarah Dean, Wenshuo Guo, Karl Krauth, Tijana Zrnic, Lydia Liu, Clara Fannjiang and Serena Wang. Many thanks to the members of the Yosef lab, including Xiuwei Zhang, David Detomaso, Jim Kaminski, Shaked Afik, Michael Cole, Allon Wagner, Zoë Steier, Carlos Buen Abad Najar, Nick Everetts, Tal Ashuach, Matt Jones, Alex Khodaverdian, Sebastian Prillo, and Galen Xing. I would like to acknowledge visiting students of the Yosef lab, for their work on the scVI team: Maxime Langevin, Edouard Mehlman, Yining Liu, Jules Samaran, Achille Nazaret, Oscar Clivio, Gabriel Misrachi, and Khalil Ouardini. Thanks to the broader member of the EECS community, including Giulia Guidi, Michael Chang, Nick Bhattacharya, Neil Thomas, Spencer Kent, and David Brookes for their friendship.

Many of the aforementioned friends have also been collaborators for the research work this thesis revolves around. In addition to them, I would like to confer distinguished MERCI to three of my closest friends and co-workers. To Adam Gayoso for leading the work on scvi-tools, and for his outstanding croissant quality classifier. To Chenling Xu for leading the work on scANVI, its application to Tabula Sapiens, and her love for cooking that reminds me of home. To Pierre Boyeau, for our work on Bayesian decision-making, and his ever-growing passion for seafood. 

Outside of the research sphere, I would like to acknowledge my dearest Bay Area friends Christelle Rohault, Antoine Dedieu, Loic Martinot, Amanda Youssef, Jerome Thai, Philippe Laban, Nina Miolane, Bertrand Travacca, Chiraz Zribi, Madisson Lips, Benjamin Fildier,  Titouan Jehl, and Paul-Armand Veillon. I am grateful to Hector Roux de B\'ezieux for our common passion for craftsmanship, especially noticeable while we built the bar at our California street house. The warmest recognition goes to Geoffrey N\'egiar, a truly inspiring friend and housemate for most of these years. Thanks for all the wonderful time spent together, crafting the finest cocktails, and cooking the most exquisite dishes in sauce. Finally, to my best friend and partner, Anna Matlin, for making the world more fun to explore and for making me a better person. Thanks for encouraging me to pursue my dreams and passions, of which this thesis is a significant part.

During the last five years, I have had many fruitful ideas while traveling. Indeed, several
key computational aspects of this thesis were sketched while visiting my hometown in
southern France. Perhaps the source for this inspiration was the many strolls through the
beautiful bauxite basin of B\'edarieux, accompanied by my parents and my brother, as we
picked up porcini, chanterelles, and saffron milk caps. Thank you for inculcating me with
the value of hard work, and perseverance, and for always believing in me. To you three, I
dedicate this thesis.
\end{acknowledgements}